What Drives State-Sponsored Violence?: Evidence from Extreme Bounds Analysis and Ensemble Learning Models

### 2018-12-04 paper-killings/manuscript.tex
add files

Four variables are robustly and positively associated with Ulfelder and Valentino's (\citeyear{ulfelder2008assessing}) indicator of government-sponsored violence. Onset and continuation of civil wars are correlated with mass killings, but only when I employ the UCDP measures of violent conflict. I find no effect for the variables compiled by the Correlates of War project or \cite{cederman2010ethnic}. Former instances of political turmoil also have a positive coefficient in our models. Moreover, countries with a previous history of riots are more prone to state violence, which suggests that government repression is path dependent \citep[e.g.,][]{gurr2000peoples,harff2003no,krain1997state,nyseth2017re}. The results also show that higher levels of ethnic diversity increase the likelihood of atrocities against civilians. Nevertheless, ethnic diversity does not pass all additional tests I implement below and the sturdiness of this finding remains open to question.

Overall, the EBA indicates two patterns in answer to our first question on the causes of mass killing. Atrocity is (1) more likely when violence is already present, reducing the costs of escalating brutality and (2) is less likely as domestic and international constraints increase, increasing the costs of escalating violence further. These patterns support the dominant opportunity narrative in the literature. However, several of the variables commonly used to proxy opportunity, such as military size or regime change, are not robust predictors of atrocity. Thus, this analysis helps clear away much of the brush around the opportunity argument.

Figure \ref{fig:drfuv} presents the ten most important predictors of state-sponsored violence in our random forest models. In general, the machine learning estimations have a good fit, with an AUC of about 0.83 in the validation sample. The algorithm confirms some of the main findings of EBA, yet they also show some interesting patterns. Only democracy and state capacity appear to be robust explanators of mass killing in both EBA and random forest models. These patterns further support the refrain that stable states tend to stay stable states.

@@

Interestingly, several variables that are not robust explanators of mass killing in the EBA, have large importance in the machine learning estimates. Variables related to characteristics of the military forces are a good example. As we see below, parametrisation and interactions likely account for this difference. The linear model imposes a parametric structure to the covariate, and the relationship between an independent variable and the response may be a nonlinear one. Also, variables can be relevant predictors only when in interaction with each other. In both cases, those relationships will be captured in the machine learning estimations but not in the extreme bounds analysis. This provides evidence that model specification is driving some of the results in the EBA.

@@

Table \ref{tab:ucdp} presents our EBA results when we restrict our analysis to only civil war years to answer Question 2. We consider three different codings of civil war: (1) the Uppsala Conflict Database Program  (\citeyear{allansson2017organized,gleditsch2002armed}), (2) the Correlates of War project \citep{sarkees2010resort}, and (3) ethnic civil war from \citet{cederman2010ethnic}. We find two important patterns. First, considering only civil war years provides a very different understanding of atrocity. Across these models, the only similarity with the full analysis is that mass killing is less likely post-Cold War. Instead, military factors, such military size and militias, and territorial war aims are the most robust predictors of atrocity once war begins. However, contrary to past expectation \citep{koren2017means}, militias have a negative impact on the likelihood of mass killings. Second, there is wide variation in which variables are robust depending on how we code civil war. Across our three codings, no variable is robust to all codings and only territorial aims and militias are robust to more than one coding. These results are concerning for scholars using correlation models, as they indicate that our understanding of atrocity, from null hypothesis testing, is largely dependent on which coding of civil war we use. For example, only the UCDP data suggests that the post-Cold War years see less barbarism than during the Cold War.

@@

When we analyse the three codings of civil war using our random forest analysis, we find further intricacies in the patterns of mass killing. First, the machine learning estimates highlight a different set of variables than the EBA when analysing the UCDP and Ethnic War data. However, the COW EBA and machine learning analyses both highlight the importance of human rights, previous riots, and the time since the state last engaged in mass killing. Thus, the COW analysis provides the most stable picture of atrocity during civil war. It again highlights the important pattern of the Conflict Trap: violence breeds violence. Second, though, the three codings of civil war each highlight a very similar set of strong predictors of atrocity during conflict. Therefore, the machine learning estimates are not as dependent on the data set employed as are the EBA results. This is good news for scholars of mass killing because it indicates that while our parametric models do not produce robust findings across different civil war data sets, our nonlinear models are able to given us a consistent and clear picture of which factors place a country at the greatest risk for atrocity during civil war.

@@

Indeed we see different patterns in both periods. In the Cold War years, Log GDP per capita has a negative impact on mass killings, while instances of previous riots increase the likelihood of state-led atrocities. The results are in line with those of the pooled model. However, mass killings seem to follow a separate logic in the post-Cold War years. Four independent variables lower the risk of mass killings: ethnic war onset, wars fought for territorial aims, coups d'état, and the share of discriminated population. I interpret the results as showing that ethnic wars are fought by groups with similar capabilities, thus large-scale, one-sided violence is relatively rare. This also explains why atrocities are more likely to occur when the share of discriminated population is not very large. The models show that territorial wars lead to more mass killings than governmental conflicts, a findings which has been previously described in the literature \citep[240]{eck2007one}. Coups d'état are correlated with fewer atrocities as well, what stands in contrast with previous research \citep[10]{wayman2010explaining}.

@@

>>>>

Four variables are robustly and positively associated with Ulfelder and Valentino's (\citeyear{ulfelder2008assessing}) indicator of government-sponsored violence. Onset and continuation of civil wars are correlated with mass killings, but only when I employ the UCDP measures of violent conflict. I find no effect for the variables compiled by the Correlates of War project or \cite{cederman2010ethnic}. Former instances of political turmoil also have a positive coefficient in the models. Moreover, countries with a previous history of riots are more prone to state violence, which suggests that government repression is path dependent \citep[e.g.,][]{gurr2000peoples,harff2003no,krain1997state,nyseth2017re}. The results also show that higher levels of ethnic diversity increase the likelihood of atrocities against civilians. Nevertheless, ethnic diversity does not pass all additional tests I implement below and the sturdiness of this finding remains open to question.

Overall, the EBA indicates two patterns in answer to my first question on the causes of mass killing. Atrocity is (1) more likely when violence is already present, reducing the costs of escalating brutality and (2) is less likely as domestic and international constraints increase, increasing the costs of escalating violence further. These patterns support the dominant opportunity narrative in the literature. However, several of the variables commonly used to proxy opportunity, such as military size or regime change, are not robust predictors of atrocity. Thus, this analysis helps clear away much of the brush around the opportunity argument.

Figure \ref{fig:drfuv} presents the ten most important predictors of state-sponsored violence in the random forest models. In general, the machine learning estimations have a good fit, with an AUC of about 0.83 in the validation sample. The algorithm confirms some of the main findings of EBA, yet they also show some interesting patterns. Only democracy and state capacity appear to be robust explanators of mass killing in both EBA and random forest models. These patterns further support the refrain that stable states tend to stay stable states.

@@

Interestingly, several variables that are not robust explanators of mass killing in the EBA, have large importance in the machine learning estimates. Variables related to characteristics of the military forces are a good example. As seen below, parametrisation and interactions likely account for this difference. The linear model imposes a parametric structure to the covariate, and the relationship between an independent variable and the response may be a nonlinear one. Also, variables can be relevant predictors only when in interaction with each other. In both cases, those relationships will be captured in the machine learning estimations but not in the extreme bounds analysis. This provides evidence that model specification is driving some of the results in the EBA.

@@

Table \ref{tab:ucdp} presents the EBA results when I restrict the analysis to only civil war years to answer Question 2. I consider three different codings of civil war: 1) the Uppsala Conflict Database Program \citeyear{allansson2017organized,gleditsch2002armed}), 2) the Correlates of War project \citep{sarkees2010resort}, and 3) ethnic civil war from \citet{cederman2010ethnic}. I find two important patterns. First, considering only civil war years provides a very different understanding of atrocity. Across these models, the only similarity with the full analysis is that mass killing is less likely post-Cold War. Instead, military factors, such military size and militias, and territorial war aims are the most robust predictors of atrocity once war begins. However, contrary to past expectation \citep{koren2017means}, militias have a negative impact on the likelihood of mass killings. Second, there is wide variation in which variables are robust depending on how scholars code civil war. Across the three codings I use here, no variable is robust to all codings and only territorial aims and militias are robust to more than one coding. These results are concerning for scholars using correlation models, as they indicate that our understanding of atrocity, from null hypothesis testing, is largely dependent on which coding of civil war researchers use. For example, only the UCDP data suggests that the post-Cold War years see less barbarism than during the Cold War.

@@

When I analyse the three codings of civil war using random forest analysis, I find further intricacies in the patterns of mass killing. First, the machine learning estimates highlight a different set of variables than the EBA when analysing the UCDP and Ethnic War data. However, the COW EBA and machine learning analyses both highlight the importance of human rights, previous riots, and the time since the state last engaged in mass killing. Thus, the COW analysis provides the most stable picture of atrocity during civil war. It again highlights the important pattern of the Conflict Trap: violence breeds violence. Second, though, the three codings of civil war each highlight a very similar set of strong predictors of atrocity during conflict. Therefore, the machine learning estimates are not as dependent on the data set employed as are the EBA results. This is good news for scholars of mass killing because it indicates that while the parametric models do not produce robust findings across different civil war data sets, the nonlinear models are able to given us a consistent and clear picture of which factors place a country at the greatest risk for atrocity during civil war.

@@

Indeed there are different patterns in both periods. In the Cold War years, Log GDP per capita has a negative impact on mass killings, while instances of previous riots increase the likelihood of state-led atrocities. The results are in line with those of the pooled model. However, mass killings seem to follow a separate logic in the post-Cold War years. Four independent variables lower the risk of mass killings: ethnic war onset, wars fought for territorial aims, coups d'état, and the share of discriminated population. I interpret the results as showing that ethnic wars are fought by groups with similar capabilities, thus large-scale, one-sided violence is relatively rare. This also explains why atrocities are more likely to occur in countries where the share of discriminated population is not very large. The models show that territorial wars lead to more mass killings than governmental conflicts, a findings which has been previously described in the literature \citep[240]{eck2007one}. Coups d'état are correlated with fewer atrocities as well, what stands in contrast with previous research \citep[10]{wayman2010explaining}.

The random forest estimations

@@

As a last robustness test for the EBA, I ran the main model with peace years only; that is, only country-years in which the UCDP, COW and Cederman et al's measures of civil conflicts had a value of zero. The results are very similar to the original model, what indicates that the difference in the results is conditional on civil war years.
